---
title: "Project 1"
author: "Antonio Jurlina"
date: "2/1/2020"
output: pdf_document
fontsize: 10pt 
geometry: margin=1in
urlcolor: blue
---

```{r setup, include=FALSE}
library(dplyr)
library(broom)
library(formattable)

directory <- paste0(here::here(), "/bootstrap")
setwd(directory)

load("income_credit_card_data.RData")

income_credit_card_data <- income_credit_card_data %>%
  mutate(age_sq = age ^ 2)

attach(income_credit_card_data)
```

### Time Series Econometrics
### Dr. Wiesen
### Project 1

**1.)** Let $x_i$ and $y_i$ (where $i = 1,2,...,N$) be iid observations. To help me prove the unbiasedness of the
OLS estimator, I used the following expression in class.

$$ \sum_{i=1}^{N} (x_i- \bar x)(y_i - \bar y) = \sum_{i=1}^{N} (x_i - \bar x)y_i $$
Prove the above expression. 

$$ \sum_{i=1}^{N} (x_i- \bar x)(y_i - \bar y) = $$
$$ \sum_{i=1}^{N} (x_i - \bar x)y_i - \sum_{i=1}^{N} (x_i - \bar x) \bar y = $$
$$ \sum_{i=1}^{N} (x_i - \bar x)y_i - \sum_{i=1}^{N} x_i \bar y + \sum_{i=1}^{N} \bar x \bar y = $$
$$ \sum_{i=1}^{N} (x_i - \bar x)y_i - N \bar x \bar y + N \bar x \bar y = $$
$$ \sum_{i=1}^{N} (x_i - \bar x)y_i $$
\newpage  

**2.)** Consider the following regression:
$$ y_i = \beta_0 + \beta_1 x_i + \varepsilon_i \hspace{0.5cm} , \hspace{0.5cm} i = 1,2,...,N $$ 
Assume that the conditional mean assumption is true. Namely, assume that $E(\varepsilon_i|x_1,x_2,...,x_n) = E(\varepsilon_i|x_i) = 0$
Also assume that the OLS estimator $\hat {\beta_1}$ is unbiased (I showed a proof of this in class).
Prove that the OLS estimator $\hat {\beta_0}$. Namely, prove that $E(\hat {\beta_0}) = \beta_0$. Hint: for the last
step of the proof use the law of iterated expectations. 

$$ \hat {\beta_0} = \bar y - \hat {\beta_1} \bar x $$ <br />  
$$ E(\hat {\beta_0}) = E(\bar y) - E(\hat {\beta_1} \bar x) $$ <br />  
$$ E(\hat {\beta_0}) = E[E(y)] - E(\hat {\beta_1} \bar x) $$ <br />  
$$ E(\hat {\beta_0}) = E[E(\beta_0 + \beta_1x_i + \varepsilon_i)] - \bar x \beta_1 $$ <br />  
$$ E(\hat {\beta_0}) = E[E(\beta_0) + E(\beta_1x_i) + E(\varepsilon_i)] - \bar x \beta_1 $$ <br />  
$$ E(\hat {\beta_0}) = E[E(\beta_0) + E(\beta_1x_i)] - \bar x \beta_1 $$ <br />  
$$ E(\hat {\beta_0}) = E(\beta_0 + \bar x \beta_1) - \bar x \beta_1 $$ <br />  
$$ E(\hat {\beta_0}) = E(\beta_0) + E(\bar x \beta_1) - \bar x \beta_1 $$ <br />  
$$ E(\hat {\beta_0}) = \beta_0 + \bar x \beta_1 - \bar x \beta_1 $$ <br />  
$$ E(\hat {\beta_0}) = \beta_0 $$ 
\newpage  

**3.)** For this problem, use the data set *Income_credit_card_data.xlsx*. The variable *AGE*
is age, the variable *INCOME* is monthly income, the variable *INCPER* is income divided by
number of dependents, and the variable *SPENDING* is average monthly credit card
expenditures.
Consider the following regression:
$$ SPENDING_i = \beta_0 + \beta_1 AGE_i + \beta_2 AGE^2_i + \beta_3 INCPER_i + \varepsilon_i \hspace{0.5cm} , \hspace{0.5cm} i = 1,2,...,N $$ 
Make the usual assumptions; the equation is true equation, the conditional mean assumption
holds, the conditional errors are iid, and the observations are iid. Use R to estimate the above
regression and report the results.  
  
```{r regression, echo = FALSE, results = 'hide'}
reg <- lm(spending ~ age + age_sq + incper) %>%
  tidy() %>%
  mutate(estimate = round(estimate, 4),
         std.error = round(std.error, 4),
         statistic = round(statistic, 4),
         p.value = round(p.value, 4))
```  
  
term      |  estimate  | standard error | t statistic | p value
--------- | ---------- | -------------- | ----------- | -----------
intercept |`r reg[1,2]`|`r reg[1,3]`    |`r reg[1,4]` |`r reg[1,5]`
age       |`r reg[2,2]`|`r reg[2,3]`    |`r reg[2,4]` |`r reg[2,5]`
age_sq    |`r reg[3,2]`|`r reg[3,3]`    |`r reg[3,4]` |`r reg[3,5]`
incper    |`r reg[4,2]`|`r reg[4,3]`    |`r reg[4,4]` |`r reg[4,5]`
  
Say you want to know at what age is SPENDING maximized. The age at which SPENDING
reaches its maximum is a function of the $\beta$ parameters, call this $\theta$ In R, use the bootstrapping
procedure to calculate an estimate of the variance of $\hat \theta$. Let the number of bootstrapped samples
be $M = 20000$. 
  
  
```{r bootstrap, echo = TRUE}
bootstrap <- function(x, M) {
  
  theta <- c(1:M)
 
  for (j in 1:M){
    data <- sample_n(x, size = nrow(x), replace = TRUE) 
    reg <- lm(data$spending ~ data$age + data$age_sq + data$incper)
    reg %>%
      summary() %>% 
      coef() %>% 
      .[2, 1] -> beta_1
    reg %>%
      summary() %>% 
      coef() %>% 
      .[3, 1] -> beta_2
    theta[j] <- -0.5 * beta_1 / beta_2
  }
  
  theta_var <- var(theta)
  
  return(theta_var)
}

theta_var <- bootstrap(x = income_credit_card_data, M = 1000)
```

The variance of $\hat \theta$ is `r theta_var` .  

\vspace*{\fill}

More at [https://github.com/antoniojurlina/time_series_econometrics](https://github.com/antoniojurlina/time_series_econometrics)  

